XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-24 talk/icooolps2011/paper.tex
reorder and generally improve things. regenerate the figure

\ifthenelse{\isundefined{\hypersetup}}{
  \usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
}{}
\hypersetup{
  pdftitle={Controlling the Tracing of an Interpreter With Hints, Part 1: Controlling the Extent of Tracing},
}

@@

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. Some of these functions contain
loops, others don't. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or reduce
the number of operations: most of these are well known compiler optimization
techniques, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}
    \item constant folding
    \item common subexpression elimination
    \item allocation removal \cite{bolz_allocation_2011}
    \item store/load propagation
    \item loop invariant code motion
\end{itemize}

@@

@@

\section{Putting Things Together}

In this section we describe a worked-out example of
a small object model for a dynamic language and how to make it efficient using
the hints described in the previous sections.

\subsection{A Simple Object Model}

To implement a dynamic language efficiently, the operations on its objects need
to be fast. Most dynamic languages have object models that are made by using
dictionaries everywhere. Let's look at an example of how the JIT can be made to
optimize such operations.

For the purpose of this section we will use a very simple and bare-bones
object model that just supports very simple classes and instances, without any
inheritance or any fancy features. The model has classes, which contain methods.
Instances have a class. Instances have their own attributes. When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' attributes are searched.

To implement this object model, we could use the following RPython code as part
of the interpreter source code:

In this straightforward implementation the methods and attributes are just
stored in dictionaries on the classes/instances. While this object model is very
simple it already contains all the hard parts of Python's object model. Both
instances and classes can have arbitrary fields, and they are changeable at
any time.  Moreover, instances can change their class after they have been
created.

When using this object model in
an interpreter, a huge amount of time will be spent doing lookups in these
dictionaries. To make the language efficient using a tracing JIT, we need to
find a way to get rid of these dictionary lookups somehow.

Let's assume we trace through code that sums three attributes, such as:

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
five calls to \texttt{dict.get}, which is slow.

@@

Therefore it makes sense to factor the layout information out of the instance
implementation into a shared object. This shared layout object is called a
\emph{map}. Maps are an old idea that comes originally from the SELF language \cite{XXX}. They are
also used by many JavaScript implementations such as V8.

The rewritten \texttt{Instance} class using maps looks like this:

@@

Instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. The maps are shared between
objects with the same layout. Therefore they have to be immutable, which means
that their \texttt{getindex} method is a pure function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{new\_map\_with\_additional\_attribute} method on the previous map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this changed instance implementation, the trace we had above changes to the
following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
has been promoted, see Figure~\ref{fig:trace2}. Operations that can be
optimized away are grayed out.

@@

Instances were optimized making the assumption that the total number of
Instance layouts is small compared to the number of instances. For classes we
will make an even stronger assumption. We simply assume that it is rare for
classes to change at all. This is not totally reasonable (sometimes classes contain
counters or similar things) but for this simple example it is good enough.

@@

Therefore, we give every class a version number, which is increased every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} looks like
in Figure~\ref{fig:version}

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
pure (because when the version number changes, the result of the call might be
different than the previous one).

@@

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{change\_method} method on the class.

@@

The first problem that needs to be solved is that Python supports (multiple)
inheritance. Therefore looking up a method in a class needs to consider the
whole method resolution order. This makes the versioning of classes more
complex. If a class is changed its version changes. At the same time, the
versions of all the classes inheriting from it need to be changed as well,
recursively. This makes class changes expensive, but they should be rare.  On the
other hand, a method lookup in a complex class hierarchy is as optimized in the
trace as in our object model here.

>>>>

@@

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or simplify
the operations in the trace. Most of these are well known compiler optimization
techniques, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}
    \item constant folding
    \item common subexpression elimination
    \item allocation removal \cite{bolz_allocation_2011}
    \item store/load propagation
    \item loop invariant code motion
\end{itemize}

@@

\subsection{Running Example}
\label{sub:running}

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other fancy features. The model has classes, which contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' methods are searched.

To implement this object model, we could use the RPython code in
Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
In this straightforward implementation the methods and attributes are just
stored in dictionaries (hash maps) on the classes and instances, respectively.
While this object model is very
simple it already contains all the hard parts of Python's object model. Both
instances and classes can have arbitrary fields, and they are changeable at
any time.  Moreover, instances can change their class after they have been
created.

When using this object model in
an interpreter, a huge amount of time will be spent doing lookups in these
dictionaries.
Let's assume we trace through code that sums three attributes, such as:

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups somehow. How to achieve this will be topic of
Section~\ref{sec:putting}.

% subsection Running Example (end)

@@

\section{Putting It All Together}

In this section we describe how the simple object model from
Section~\ref{sub:running} can be made efficient using the hints described in the
previous the section. The object model there is typical for many current
dynamic languages (such as Python, Ruby and JavaScript) as it relies heavily on
hash-maps to implement its objects.

@@

Therefore it makes sense to factor the layout information out of the instance
implementation into a shared object, called the \emph{map}. Maps are a well-known
technique to efficiently implement instances and come from the SELF project
\cite{XXX}. They are also used by many JavaScript implementations such as V8.
The rewritten \texttt{Instance} class using maps can be seen in
Figure~\ref{fig:maps}.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. The maps are shared between
objects with the same layout. Therefore they have to be immutable, which means
that their \texttt{getindex} method is a pure function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map (which is also pure). Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this changed instance implementation, the trace we had above changes to the
following that of see Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
memory address of the \texttt{Map} instance that has been promoted. Operations
that can be optimized away are grayed out.

@@

Instances were optimized making the assumption that the total number of
different instance layouts is small compared to the number of instances. For classes we
will make an even stronger assumption. We simply assume that it is rare for
classes to change at all. This is not totally reasonable (sometimes classes contain
counters or similar things) but for this simple example it is good enough.

@@

Therefore, we give every class a version number, which is changed every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
pure, because when the version number changes, the result of the call might be
different than the previous one.

@@

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if the trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{change\_method} method on the class.

@@

The first problem that needs to be solved is that Python supports (multiple)
inheritance. Therefore looking up a method in a class needs to consider all the
classes in the
whole method resolution order. This makes the versioning of classes more
complex. If a class is changed its version changes. At the same time, the
versions of all the classes inheriting from it need to be changed as well,
recursively. This makes class changes expensive, but they should be rare.  On the
other hand, a method lookup in a complex class hierarchy is as optimized in the
trace as in our object model here.
